---
title: "RNA-seq analysis in R"
author: "Stephane Ballereau, Mark Dunning, Oscar Rueda, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_notebook:
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    toc_float: yes
layout: page
subtitle: Pre-processsing RNA-seq data - Challenge Solutions
---

## Data 

```{r setup, message = FALSE}
library(DESeq2)
library(gplots)
library(RColorBrewer)
library(limma)
library(tidyverse)
```

```{r prepareData, message=FALSE}
# Read the sample information into R
sampleinfo <- read.delim("../data/SampleInfo.Corrected.txt", 
                         stringsAsFactors = F)
# Read the data into R
seqdata <- read.delim("../data/GSE60450_Lactation.featureCounts", 
                      stringsAsFactors = F,
                      comment = "#")
# Remove first two columns from seqdata
countdata <- as.data.frame(seqdata) %>% 
    column_to_rownames("Geneid") %>% # turn the geneid column into rownames
    rename_all(str_remove, ".bam") %>% # remove the ".bam" from the column names
    select(sampleinfo$Sample) %>% # keep sample columns using sampleinfo
    as.matrix()
# filter data
keep <- rowSums(countdata) > 5
countdata <- countdata[keep,]
# rlogcounts
rlogcounts <- rlog(countdata)
# We estimate the variance for each row in the logcounts matrix
countVar <- apply(rlogcounts, 1, var)
# DGE list
design <- as.formula(~ CellType)
# create the DESeqDataSet object
ddsObj <- DESeqDataSetFromMatrix(countData = countdata,
                              colData = sampleinfo,
                              design = design)
ddsObj <- estimateSizeFactors(ddsObj)
# normalise dcounts
logcounts <- log2(countdata + 1)
normalizedCounts <- counts(ddsObj, normalized=TRUE) 
logNormalizedCounts <- log2(normalizedCounts + 1)
```

> ### Challenge 1
>
> 1. Use the `DESeq2` function `rlog` to transform the count data. This function
> also normalises for library size.
> 2. Plot the count distribution boxplots with this data
> How has this effected the count distributions?

```{r}
rlogcounts <- rlog(countdata)

statusCol <- as.numeric(factor(sampleinfo$Status)) + 1
# Check distributions of samples using boxplots
boxplot(rlogcounts, 
        xlab="", 
        ylab="Log2(Counts)",
        las=2,
        col=statusCol)
# Let's add a blue horizontal line that corresponds to the median logCPM
abline(h=median(as.matrix(logcounts)), col="blue")

```

> ## Challenge 2 {.challenge}
>
> Redo the heatmap using the top 500 LEAST variable genes.
> Change the colour scheme to "PiYG" and redo the heatmap. Try `?RColorBrewer` and see what other colour schemes are available.
> Change the sample names to `group` using the `labCol` argument
> Remove the gene names from the righthand side of the plot using `labRow`

**Solution**
```{r solutionChallenge2, fig.height=15, fig.width=10}

# Get the gene names for the top 500 least variable genes
lowVar <- order(countVar)[1:500]
# Subset logcounts matrix
hmData <- rlogcounts[lowVar,]

## Get some nicer colours
mypalette <- brewer.pal(11,"PiYG")
## http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
morecols <- colorRampPalette(mypalette)
# Set up colour vector for celltype variable
col.cell <- c("purple","orange")[sampleinfo$CellType]

# Plot the heatmap
heatmap.2(hmData, 
          col=rev(morecols(50)),
          trace="none", 
          main="Top 500 most variable genes across samples",
          ColSideColors=col.cell,scale="row",
          labCol=sampleinfo$Group, 
          labRow = NA)

```

> ### Challenge 3
>
> Plot the biased and unbiased MA plots both samples side by side to see the 
> before and after normalisation.
>

```{r solutionChallenge3, echo=FALSE, fig.height=10, fig.width=10}
par(mfrow=c(2,2))
plotMA(logcounts, array = 7, main="MCL1.LA - Raw")
abline(h=0,col="grey")
plotMA(logNormalizedCounts, array = 7, main="MCL1.LA - Normalised")
abline(h=0,col="grey")
plotMA(logcounts, array = 11, main="MCL1.LE - Raw")
abline(h=0,col="grey")
plotMA(logNormalizedCounts, array = 11, main="MCL1.LE - Normalised")
abline(h=0,col="grey")
```