---
title: "RNA-seq Analysis in R"
subtitle: "Annotation and Visualisation of RNA-seq results - Solutions"
author: "Stephane Ballereau, Dominique-Laurent Couturier, Mark Dunning, Abbi Edwards, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
minutes: 300
layout: page
---

```{r setup, include=FALSE}
library(biomaRt)
library(GenomicFeatures)
library(DESeq2)
library(tidyverse)
```

```{r prepareData, echo=FALSE, message=FALSE, warning=FALSE}
# load data
load("../Robjects/DE.RData")
## set up connection to ensembl database
ensembl <- useMart("ENSEMBL_MART_ENSEMBL")
# list the available datasets (species)
listDatasets(ensembl) %>% 
    filter(str_detect(description, "Mouse"))
# specify a data set to use
ensembl <- useDataset("mmusculus_gene_ensembl", mart=ensembl)
# annotations
load("../Robjects/Ensembl_annotations.RData")
load("../Robjects/Full_annotation.RData")

annotLvV <- as.data.frame(resLvV) %>% 
    rownames_to_column("GeneID") %>% 
    left_join(ensemblAnnot, "GeneID") %>% 
    rename(logFC=log2FoldChange, FDR=padj)
ddsShrink <- lfcShrink(ddsObj, coef="Status_lactate_vs_virgin")
shrinkLvV <- as.data.frame(ddsShrink) %>%
    rownames_to_column("GeneID") %>% 
    left_join(ensemblAnnot, "GeneID") %>% 
    rename(logFC=log2FoldChange, FDR=padj)
```

```{r prepareData3, echo=FALSE, message=FALSE, warning=FALSE}
# txdb object
txMm <- makeTxDbFromGFF("../data/counts/Mus_musculus.GRCm38.97.gtf.gz")
```


### Challenge 1

That was just 1000 genes. We need annotations for the entire results table.
Also, there may be some other interesting columns in BioMart that we wish to
retrieve.  

> (a) Search the attributes and add the following to our list of attributes:  
>       (i) The gene description   
>       (ii) The gene biotype  
> (b) Query BioMart using all of the genes in our results table (`resLvV`)  
> (c) How many Ensembl genes have multipe Entrez IDs associated with them?  

```{r solutionChallenge1code, eval=FALSE, message=FALSE, warnings=FALSE}
# Set the filter type and values
ourFilterType <- "ensembl_gene_id"

# set the values for the filter
filterValues <- rownames(resLvV)

# check the available "attributes" - things you can retreive
listAttributes(ensembl) %>%
    filter(str_detect(name, "description"))

listAttributes(ensembl) %>%
    filter(str_detect(name, "biotype"))
```

```{r solutionChallenge1, echo=FALSE, message=FALSE, warnings=FALSE}
# Set the filter type and values
ourFilterType <- "ensembl_gene_id"

# set the values for the filter
filterValues <- rownames(resLvV)

# check the available "attributes" - things you can retreive
listAttributes(ensembl) %>%
    filter(str_detect(name, "description"))

listAttributes(ensembl) %>%
    filter(str_detect(name, "biotype"))

```

```{r solutionChallenge1acode, eval=FALSE, message=FALSE, warnings=FALSE}
# set attributes
attributeNames <- c('ensembl_gene_id',
                    'entrezgene_id',
                    'external_gene_name',
                    'description',
                    'gene_biotype')

# run the query
annot <- getBM(attributes=attributeNames,
               filters = ourFilterType,
               values = filterValues,
               mart = ensembl)

# count duplicate ids
annot %>%
  add_count(ensembl_gene_id) %>% 
  filter(n>1) %>% 
  distinct(ensembl_gene_id) %>% 
  nrow()
```

```{r solutionChallenge1a, echo=FALSE, message=FALSE, warnings=FALSE}
# duplicate ids
annot %>%
  add_count(ensembl_gene_id) %>% 
  filter(n>1) %>% 
  distinct(ensembl_gene_id) %>% 
  nrow()
```

### Challenge 2 {.challenge}

Use the log2 fold change (`logFC`) on the x-axis, and use `-log10(FDR)` on the
y-axis. (This >`-log10` transformation is commonly used for p-values as it
means that more significant genes have a higher scale).

> (a) Create a column of -log10(FDR) values
>
> (b) Create a plot with points coloured by if FDR < 0.05


```{r volcanoPlot, fig.height=5, fig.width=7}
# first remove the filtered genes (FDR=NA) and create a -log10(FDR) column
filtTab <- shrinkLvV %>% 
    filter(!is.na(FDR)) %>% 
    mutate(`-log10(FDR)` = -log10(FDR))

ggplot(filtTab, aes(x = logFC, y=`-log10(FDR)`)) + 
    geom_point(aes(colour=FDR < 0.05), size=1)
```


### Challenge 3 - In Supplementary Materials

Use the txMm to retrieve the exon coordinates for the genes: 

> * `ENSMUSG00000021604`  
> * `ENSMUSG00000022146`  
> * `ENSMUSG00000040118`   


```{r solutionChallenge3, warning=FALSE, message=FALSE}
keyList <- c("ENSMUSG00000021604", "ENSMUSG00000022146", "ENSMUSG00000040118")
AnnotationDbi::select(txMm, 
       keys=keyList,
       keytype = "GENEID",
       columns=c("TXNAME", "TXCHROM", "TXSTART", "TXEND", "TXSTRAND", "TXTYPE")
      )
```
