---
layout: page
title: R for RNAseq analysis
subtitle: Introduction to RStudio
minutes: 45
---

```{r, include=FALSE}
source("tools/chunk-options.R")
```

> ## Learning Objectives {.objectives}
>
> * To gain familiarity with the various panes in the RStudio IDE
> * To gain familiarity with the buttons, short cuts and options in the RStudio IDE
> * To be able to manage your workspace in an interactive R session
>

## Introduction to RStudio

Welcome to the R portion of the Software Carpentry workshop.

Throughout this lesson, we're going to teach you some of the fundamentals of
the R language as well as some best practices for organising code for
scientific projects that will make your life easier.

We'll be using RStudio: a free, open source R integrated development
environment. It provides a built in editor, works on all platforms (including
on servers) and provides many advantages such as integration with version
control and project management.

**Basic layout**

When you first open RStudio, you will be greeted by three panels:

  * The interactive R console (entire left)
  * Environment/History (tabbed in upper right)
  * Files/Plots/Packages/Help/Viewer (tabbed in lower right)

Once you open files, such as R scripts, an editor panel will also open
in the top left.

## Work flow within RStudio
There are two main ways one can work within RStudio.

1. Test and play within the interactive R console then copy code into
a .R file to run later.
   *  This works well when doing small tests and initially starting off.
   *  It quickly becomes laborious
2. Start writing in an .R file and use RStudio's command / short cut
to push current line, selected lines or modified lines to the
interactive R console.
   * This is a great way to start; all your code is saved for later
   * You will be able to run the file you create from within RStudio
   or using R's `source()`  function.

> ## Tip: Running segments of your code {.callout}
>
> RStudio offers you great flexibility in running code from within the editor
> window. There are buttons, menu choices, and keyboard shortcuts. To run the
> current line, you can 1. click on the `Run` button just above the editor panel,
> or 2. select "Run Lines" from the "Code" menu, or 3. hit Ctrl-Enter in Windows
> or Linux or Command-Enter on OS X. (This shortcut can also be seen by hovering
> the mouse over the button). To run a block of code, select it and then `Run`.
> If you have modified a line of code within a block of code you have just run,
> there is no need to reselct the section and `Run`, you can use the next button
> along, `Re-run the previous region`. This will run the previous code block
> inculding the modifications you have made.

## Your RStudio environment

Much of your time in R will be spent in the R interactive
console. This is where you will run all of your code, and can be a
useful environment to try out ideas before adding them to an R script
file. This console in RStudio is the same as the one you would get if
you just typed in `R` in your commandline environment.

The first thing you will see in the R interactive session is a bunch
of information, followed by a ">" and a blinking cursor. In many ways
this is similar to the shell environment you learned about during the
shell lessons: it operates on the same idea of a "Read, evaluate,
print loop": you type in commands, R tries to execute them, and then
returns a result.

## Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1 + 100
```

And R will print out the answer, with a preceding "[1]". Don't worry about this
for now, we'll explain that later. For now think of it as indicating ouput.

Just like bash, if you type in an incomplete command, R will wait for you to
complete it:

~~~ {.r}
> 1 +
~~~

~~~ {.output}
+
~~~

Any time you hit return and the R session shows a "+" instead of a ">", it
means it's waiting for you to complete the command. If you want to cancel
a command you can simply hit "Esc" and RStudio will give you back the ">"
prompt.

> ## Tip: Cancelling commands {.callout}
>
> If you're using R from the commandline instead of from within RStudio,
> you need to use `Ctrl+C` instead of `Esc` to cancel the command. This
> applies to Mac users as well!
>
> Cancelling a command isn't just useful for killing incomplete commands:
> you can also use it to tell R to stop running code (for example if its
> taking much longer than you expect), or to get rid of the code you're
> currently writing.
>

## Vectorization

R is *vectorized*, meaning that
variables and functions can have vectors as values. For example

```{r}
1:5
2^(1:5)
x <- 1:5
2^x
```

This is incredibly powerful; we will discuss this further in an
upcoming lesson.


## Managing your environment

There are a few useful commands you can use to interact with the R session.

`ls` will list all of the variables and functions stored in the global environment
(your working R session):

```{r}
ls()
```

> ## Tip: hidden objects {.callout}
>
> Just like in the shell, `ls` will hide any variables or functions starting
> with a "." by default. To list all objects, type `ls(all.names=TRUE)`
> instead
>

Note here that we didn't given any arguments to `ls`, but we still
needed to give the parentheses to tell R to call the function.

If we type `ls` by itself, R will print out the source code for that function!

```{r}
ls
```

You can use `rm` to delete objects you no longer need:

```{r}
rm(x)
```

If you have lots of things in your environment and want to delete all of them,
you can pass the results of `ls` to the `rm` function:

```{r}
rm(list = ls())
```

In this case we've combined the two. Just like the order of operations, anything
inside the innermost parentheses is evaluated first, and so on.

In this case we've specified that the results of `ls` should be used for the
`list` argument in `rm`. When assigning values to arguments by name, you *must*
use the `=` operator!!

If instead we use `<-`, there will be unintended side effects, or you may just
get an error message:

```{r, error=TRUE}
rm(list <- ls())
```

> ## Tip: Warnings vs. Errors {.callout}
>
> Pay attention when R does something unexpected! Errors, like above,
> are thrown when R cannot proceed with a calculation. Warnings on the
> other hand usually mean that the function has run, but it probably
> hasn't worked as expected.
>
> In both cases, the message that R prints out usually give you clues
> how to fix a problem.
>


## R Packages

It is possible to add functions to R by writing a package, or by
obtaining a package written by someone else. As of this writing, there
are over 7,000 packages available on CRAN (the comprehensive R archive
network). R and RStudio have functionality for managing packages:

* You can see what packages are installed by typing
  `installed.packages()`
* You can install packages by typing `install.packages("packagename")`,
  where `packagename` is the package name, in quotes.
* You can update installed packages by typing `update.packages()`
* You can remove a package with `remove.packages("packagename")`
* You can make a package available for use with `library(packagename)`
