---
title: "RNA-seq analysis in R"
author: "Stephane Ballereau, Mark Dunning, Oscar Rueda, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_notebook:
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    toc_float: yes
minutes: 300
layout: page
subtitle: Gene Set Testing for RNA-seq
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Original Authors: Belinda Phipson, Anna Trigos, Matt Ritchie, Maria Doyle, Harriet Dashnow, Charity Law**
Based on the course [RNAseq analysis in R](http://combine-australia.github.io/2016-05-11-RNAseq/) delivered on May 11/12th 2016


```{r}
library(edgeR)
load("Robjects/DE.Rdata")
```

# Testing relative to a threshold (TREAT)

When there is a lot of differential expression, sometimes we may want to cut-off on a fold change threshold as well as a p-value threshold so that we follow up on the most biologically significant genes. However, it is not recommended to simply rank by p-value and then discard genes with small logFC's, as this has been shown to increase the false discovery rate. In other words, you are not controlling the false discovery rate at 5\% anymore.

The test performed above tests the null hypothesis that gene expression is the same in classes compared in the contrast of interest, i.e. that fold change is 1 (and log2(FC) is 0). Rather than ranking genes per p-values and then filter on logFC, one should instead test the null hypothesis that the difference in level of expression between classes in the contrast is lower than a given threshold.

See ["Testing significance relative to a fold-change threshold is a TREAT"](https://academic.oup.com/bioinformatics/article/25/6/765/251641/Testing-significance-relative-to-a-fold-change)

## Fit the linear model

Remember how the linear model was fit in the previous section:

```{r}
fit <- glmFit(dgeObj, design)
```

Let's fit the model again:

```{r}
## Read the counts from the downloaded data
seqdata <- read.delim("data/GSE60450_Lactation-GenewiseCounts.txt", stringsAsFactors = FALSE)
#
# Remove first two columns from seqdata

countdata <- seqdata[,-(1:2)]

# Store EntrezGeneID as rownames
rownames(countdata) <- seqdata[,1]
countdata
colnames(countdata) <- substr(colnames(countdata), 1, 7)
countdata

## Calculate the Counts Per Million measure
myCPM <- cpm(countdata)

## Identify genes with at least 0.5 cpm in at least 2 samples
thresh <- myCPM > 0.5
keep <- rowSums(thresh) >= 2

# Subset the rows of countdata to keep the more highly expressed genes
counts.keep <- countdata[keep,]

## Convert to an edgeR object
dgeObj <- DGEList(counts.keep)

## Perform TMM normalisation
dgeObj <- calcNormFactors(dgeObj)

# Estimate dispersion:
dgeObj <- estimateCommonDisp(dgeObj)
dgeObj <- estimateGLMTrendedDisp(dgeObj)
dgeObj <- estimateTagwiseDisp(dgeObj)
plotBCV(dgeObj)

# Create design matrix:
# Obtain sample information
sampleinfo <- read.delim("data/SampleInfo_Corrected.txt")
group <- paste(sampleinfo$CellType,sampleinfo$Status,sep=".")
group
# Create the two variables
group <- as.character(group)
type <- sapply(strsplit(group, ".", fixed=T), function(x) x[1])
status <- sapply(strsplit(group, ".", fixed=T), function(x) x[2])
ftable(type, status, exclude=c())

# Specify a design matrix without an intercept term
design <- model.matrix(~ type + status)
design
#colnames(design)

# Fit model:
fit <- glmFit(dgeObj, design)
names(fit)
head(coef(fit))
```

Conduct likelihood ratio tests for the luminal vs basal contrast and show the top genes:

```{r}
# Conduct likelihood ratio tests for luminal vs basal and show the top genes:
# Remember that contrast names are kept in
# - the coef(), see head(coef(fit)),
# - the design matrix, see colnames(design)

lrt.BvsL <- glmLRT(fit, coef=2) 
#class((lrt.BvsL)) # "DGELRT"

# Show top genes:
topTags(lrt.BvsL)

# Access test results kept in the 'table' slot:
head(lrt.BvsL$table)
```

Check counts per million for the most significant genes:

```{r}
# Check counts per million for the most significant genes:

# Order genes by increasing order of p-value
o <- order(lrt.BvsL$table$PValue)

# Show counts per million for the top 10 genes for all samples
round(cpm(dgeObj)[o[1:10],])
# Remember the 'sampleinfo' dataframe keeps samples description

# Maybe plot counts per million for the top gene
barplot(cpm(dgeObj)[o[1],])
```

## MD plot

Create mean-difference plot to show log fold changes (differences) versus average log values (means):

```{r}
# Create mean-difference ('MD') plot to show log fold changes (differences) versus average log values (means)
# Differentially exressed genes (DEGs) may be color-coded using the 'status' argument.
# Let's color genes according to the outcome of the LRT test.

# Classify the differential expression statistics as up, down or not significant using decideTestsDGE()
# ?decideTestsDGE
tmpDec <- decideTestsDGE(lrt.BvsL, lfc=1)

# Show the number of genes in each of the three classes:
summary(decideTestsDGE(lrt.BvsL, lfc=1))

# Create MD plot with extra blue lines to show log fold-change threshold:
edgeR::plotMD.DGELRT(lrt.BvsL, status=tmpDec); abline(h=c(-1, 1), col="blue")
```

Mind that glmLRT() called without 'coef' value uses the last contrast in the design matrix

```{r}
# glmLRT() called without 'coef' value uses the last contrast in the design matrix

# Conduct likelihood ratio tests fo all contrasts at once
lrt <- glmLRT(fit)
# class(lrt) #  "DGELRT"
lrt$comparison
# colnames(design)

topTags(lrt)
decideTestsDGE(lrt)
summary(decideTestsDGE(lrt, lfc=1))
```

## Testing relative to a threshold (TREAT) with edgeR

Read the manual on glmTreat() to "Conduct genewise statistical tests for a given coefficient or contrast relative to a specified fold-change threshold."

"‘glmTreat’ implements a test for differential expression relative
     to a minimum required fold-change threshold. Instead of testing
     for genes which have log-fold-changes different from zero, it
     tests whether the log2-fold-change is greater than ‘lfc’ in
     absolute value. ‘glmTreat’ is analogous to the TREAT approach
     developed by McCarthy and Smyth (2009) for microarrays."
"‘glmTreat’ detects whether ‘glmfit’ was produced by ‘glmFit’ or
     ‘glmQLFit’. In the former case, it conducts a modified likelihood
     ratio test (LRT) against the fold-change threshold. In the latter
     case, it conducts a quasi-likelihood (QL) F-test against the
     threshold."
     
```{r}
?glmTreat
```

We will use the "output from ‘glmFit’" created above.

### Test for abs(logFC) > 0

Let's test for abs(logFC) > 0:

```{r}
# > colnames(design)
#[1] "(Intercept)"    "typeluminal"    "statuspregnant" "statusvirgin"  

res.treat.tlum.fc0 <- glmTreat(fit, "typeluminal", contrast = NULL, lfc = 0, null = "interval")
topTags(res.treat.tlum.fc0)
#coef(res.treat.tlum.fc0)
#class(res.treat.tlum.fc0$table)
#colnames(res.treat.tlum.fc0$table)
head(res.treat.tlum.fc0$table)
```

Number of genes either up- or down-regulated, or not showing significant difference in expression between the two groups:

```{r}
summary(decideTestsDGE(res.treat.tlum.fc0))
```

Draw MD plot: 

```{r}
tmpDec.fc0 <- decideTestsDGE(res.treat.tlum.fc0)
edgeR::plotMD.DGELRT(res.treat.tlum.fc0, status=tmpDec.fc0)
abline(h=c(-1, 1), col="blue")
```

### Test for abs(logFC) > 1

Let's test for abs(logFC) > 1:

```{r}
# > colnames(design)
#[1] "(Intercept)"    "typeluminal"    "statuspregnant" "statusvirgin"  

res.treat.tlum.fc1 <- glmTreat(fit, "typeluminal", contrast = NULL, lfc = 1, null = "interval")

# Glance:
topTags(res.treat.tlum.fc1)

# Number of genes either up- or down-regulated, or not showing significant difference in expression between the two groups:
tmpDec.fc1 <- decideTestsDGE(res.treat.tlum.fc1)
summary(tmpDec.fc1)

# Draw MD plot: 
edgeR::plotMD.DGELRT(res.treat.tlum.fc1, status=tmpDec.fc1)
abline(h=c(-1, 1), col="blue")
```

### Visualise effect of TREAT on DEG list

You may have noticed that fewer genes are highlighted in the MAplot for 'abs(logFC) > 1'  than for 'abs(logFC) > 0'.

Let's identify genes flagged as DEG with 'abs(logFC) > 0' but not 'abs(logFC) > 1'.

```{r}

# compare decisions between the two tests:

ftable(as.vector(tmpDec.fc0), as.vector(tmpDec.fc1), exclude=c())

# Identify gene 'in the abs-FC>0 set but not in the abs-FC>1 set':

tmpDec.inFc0outFc1 <- as.logical(tmpDec.fc0) & ! as.logical(tmpDec.fc1)
table(tmpDec.inFc0outFc1)

edgeR::plotMD.DGELRT(res.treat.tlum.fc1, status=tmpDec.inFc0outFc1)
abline(h=c(-1, 1), col="blue")


```

> ## Challenge {.challenge}
>
> Change the cut-off so that we are interested in genes that change at least 50\% on the fold change scale.
>
> HINT: what is the corresponding logFC value of 50\% fold change? Assume basal.pregnant is 50\% higher than basal.lactate
>

# Gene Set Testing

Sometimes there is quite a long list of differentially expressed genes to interpret after a differential expression analysis, and it is usually infeasible to go through the list one gene at a time trying to understand its biological function. A common downstream procedure is gene set testing, which aims to understand which pathways/gene networks the differentially expressed genes are implicated in.

There are a number of different ways to go about testing for enrichment of biological pathways, and the test you choose usually depends on the question you're asking. There are two kinds of tests: self-contained and competitive gene set tests.

Self-contained tests, which include the `ROAST` procedure, ask the question "Are the genes in the set/pathway differentially expressed as a whole?"

Competitive gene set tests, such as those implemented in `goana` and `camera` ask the question whether the differentially expressed genes tend to be over-represented in the gene set, compared to all the other genes in the experiment.

These different questions use different statistical methodology.

## Gene Set Testing - competitive gene set tests

### goseq analysis

From the goseq vignette: "goseq first needs to quantify the length bias present in the dataset under consideration.  This is done by calculating a Probability Weighting Function or PWF which can be thought of as a function which gives the probability that a gene will be differentially expressed (DE), based on its length alone.  The PWF is calculated by fitting a monotonic spline to the binary data series of differential expression (1=DE, 0=Not DE) as a function of gene length.  The PWF is used to weight the chance of selecting each gene when forming a null distribution for GO category membership.  The fact that the PWF is calculated directly from the dataset under consideration makes this approach robust, only correcting for the length bias present in the data."

"GO analysis of RNA-seq data requires the use of random sampling in order to generate a suitable null distribution for GO category membership and calculate each categories significance for over representation amongst DE genes. ... In  most  cases,  the  Wallenius
distribution can be used to approximate the true null distribution, without any significant loss in accuracy.  The goseq package implements this approximation as its default option"

Create list of DEGs:

```{r}
results <- as.data.frame(topTags(lrt.BvsL,n = Inf))
results
genes <- results$FDR < 0.01
names(genes) <- rownames(results)
```

Fit the Probability Weighting Function (PWF)

```{r}
library(goseq)

supportedGeneIDs()
supportedGenomes()

if(FALSE)
{
TxDb.Mmusculus.UCSC.mm10.knownGene

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
		GRList <- transcriptsBy(txdb, by = "gene")
		tmp <- lapply(GRList, width) # slow
		#tmp2 <- lapply(tmp,max)
		#tmp2 <- lapply(tmp,median) # list
		tmp2 <- unlist(lapply(tmp,median)) # vector

		print("Got gene length: transcript median length.")
		print(summary(tmp2))
		print(head(tmp2))
		# Make df:
		geneLengthMedianDf <- as.data.frame(tmp2)
		colnames(geneLengthMedianDf) = "medianLength"
		geneLengthMedianDf$geneId <- rownames(geneLengthMedianDf)
		#save(geneLengthMedianDf, file=tmpFileName.med)



			# read DEG table in:
			allDf <- read.table(tmpFileName, header=T, stringsAsFactor=F, sep=",")
			allDf <- results
			print(dim(allDf))
			print(head(allDf))
#next
			# remember that table aso keep location and thus that length can be derived
			# but that would be the maximum transcript length, not the median length 

			# mind that some genes have NA padj

			# genes: boolean vector
			
			# remove gene with NA for padj
			#tmpBoolNonNa <- !is.na(allDf$padj)
			#allDfToUse <- allDf[tmpBoolNonNa,]

			allDfToUse <- allDf
			
			# identify DEG, using lfc and padj:
			lfcthd <- 0
			fdrthd <- 0.05
			# make boolean into 0 for F and 1 for T, to get tutorial 'genes'
      tmpBoolLfc <- abs(allDfToUse$logFC) > lfcthd
			#tmpBoolPadj <- allDfToUse$padj<padjthd
      tmpBoolFdr <- allDfToUse$FDR<fdrthd

			#tmpBoolDeg <- tmpBoolLfc & tmpBoolPadj
			tmpBoolDeg <- tmpBoolLfc & tmpBoolFdr
      table(tmpBoolDeg)
			
			# flag DEGs,
			#boolDeg <- as.numeric(allDfToUse$padj<0.05) # wo lfc filter
			boolDeg <- as.numeric(tmpBoolDeg) # wi lfc filter

			names(boolDeg) <- rownames(allDfToUse) 
			#print(table(boolDeg))
			#print(head(boolDeg))

			# make df
			boolDegDf <- as.data.frame(boolDeg)
			boolDegDf$geneId <- rownames(boolDegDf)
			print(head(boolDegDf))

			#boolDegDf2 <- merge(boolDegDf, geneLengthMedianDf, by="ensGeneId", all.x=T, sort=F)
			boolDegDf2 <- merge(boolDegDf, geneLengthMedianDf, by="geneId", all=F, sort=F)
			print(head(boolDegDf2))

			#  Fitting the Probability Weighting Function (PWF)
			#pwf=nullp(boolDegDf2$boolDeg,"hg38","ensGene", bias.data=boolDegDf2$medianLength)
			boolDegVec <- boolDegDf2$boolDeg
			names(boolDegVec) <- boolDegDf2$geneId
			pwf=nullp(boolDegVec,"mm10","knownGene", bias.data=boolDegDf2$medianLength)
			print(head(pwf))
}

pwf <- nullp(genes, "mm10","knownGene")
```

Conduct gene set enrichment analysis

```{r}
go.results <- goseq(pwf, "mm10","knownGene")
go.results
```

### fgsea analysis

```{r}
library(fgsea)
```

Loading ranks:

```{r}
ranks <- results$logFC
names(ranks) <- rownames(results)
```

Load the pathways

```{r}
load("data/mouse_H_v5.rdata")
pathways <- Mm.H
```

Conduct analysis

```{r}
fgseaRes <- fgsea(pathways, ranks, minSize=15,maxSize = 500,nperm=1000)
fgseaRes
```

Glance at results

```{r}
head(fgseaRes[order(pval), ])
```

Plot outcome for the 'HALLMARK_MYOGENESIS' pathway

```{r}
plotEnrichment(pathways[["HALLMARK_MYOGENESIS"]],
               ranks)
```

Select top pathways and plot outcome for all these

```{r}
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(pathways[topPathways], ranks, fgseaRes, 
              gseaParam = 0.5)
```

### CAMERA gene set testing using the Broad's curated gene sets

Other databases of gene sets that are available come from the Broad Institute's Molecular Signatures Database ([MSigDB](http://software.broadinstitute.org/gsea/msigdb)). CAMERA is good option for testing a very large number of gene sets such as the MSigDB sets, as it is very fast. CAMERA is known as a competitive gene set test, however it has the advantage that it can take into account inter-gene correlation within each gene set [@wu2012camera]. It also works seemlessly with a `voom` object, taking into account the mean-variance relationship in RNA-Seq data.

Here we will be using the C2 gene sets for mouse, available as .rdata files from the WEHI bioinformatics page [http://bioinf.wehi.edu.au/software/MSigDB/index.html](http://bioinf.wehi.edu.au/software/MSigDB/index.html). The C2 gene sets contain 4725 curated gene sets collected from a variety of places: BioCarta, KEGG, Pathway Interaction Database, Reactome as well as some published studies.It doesn't include GO terms.

```{r}

# ?camera.DGEList

# Load in the mouse c2 gene sets
# The R object is called Mm.c2
load("data/mouse_c2_v5.rdata")
# Have a look at the first few gene sets
names(Mm.c2)[1:5]
# Number of gene sets in C2
length(Mm.c2)
```

The gene identifiers are Entrez Gene ID, which is the same as the rownames of our voom object. We need to map the Entrez gene ids between the list of gene sets and our voom object. We can do this using the `ids2indices` function.

```{r}
c2.ind <- ids2indices(Mm.c2, rownames(dgeObj$counts))
```

CAMERA takes as input the voom object `v`, the indexed list of gene sets `c2.ind`, the design matrix, the contrast being tested, as well as some other arguments. By default, CAMERA can estimate the correlation for each gene set separately. However, in practise, it works well to set a small inter-gene correlation of about 0.05 using the `inter.gene.cor` argument.

```{r}
gst.camera <- camera.DGEList(dgeObj,index=c2.ind,design=design,contrast =3,inter.gene.cor=0.05)
```

CAMERA outputs a dataframe of the resulting statistics, with each row denoting a different gene set. The output is ordered by p-value so that the most significant should be at the top. Let's look at the top 5 gene sets:

```{r}
gst.camera[1:5,]
```

The total number of significant gene sets at 5\% FDR is

```{r}
table(gst.camera$FDR < 0.05)
```

You can write out the camera results to a csv file to open in excel.

```{r}
write.csv(gst.camera,file="gst_BPregVsLac.csv")
```

> ## Challenge {.challenge}
>
> 1. Run `camera` on the second contrast in the contrast matrix.
> 1. Run `camera` on a different set of MSigDB gene sets, the hallmark datasets, `mouse_H_v5.rdata`.
> You will need to load in the hallmark gene sets, and the object will be called `Mm.H` in R.
>

## Gene Set Testing - self-contained gene set tests

### ROAST gene set testing

ROAST is an example of a self-contained gene set test [@wu2010roast]. It asks the question, "Do the genes in my set tend to be differentially expressed between my conditions of interest?". ROAST doesn't care about what the other genes in the experiment are doing, which is different to `camera` and `goana`. ROAST is a good option for when you're interested in a specific set, or a few sets. It is not really used to test thousands of sets at one time.

From the Hallmark gene sets, two MYC pathways were most significant.

```{r}
H.camera[1:10,]
```

Let's see if there are any MYC signalling pathways in MsigDB C2 collection. We can do this with the `grep` command on the names of the gene sets.

```{r}
grep("MYC_",names(c2.ind))
# Let's save these so that we can subset c2.ind to test all gene sets with MYC in the name
myc <- grep("MYC_",names(c2.ind))
# What are these pathways called?
names(c2.ind)[myc]
```

Let's use ROAST to see if these MYC related gene sets tend to be differentially expressed. Note that the syntax for `camera` and `roast` is almost identical.

```{r}
myc.rst <- roast(dgeObj,index=c2.ind[myc],design=design,contrast=2,nrot=999)
myc.rst[1:15,]
```

Each row corresponds to a single gene set. The NGenes column gives the number of genes in each set. The PropDown and PropUp columns contain the proportions of genes in the set that are down- and up-regulated, respectively, with absolute fold changes greater than 2. The net direction of change is determined from the significance of changes in each direction, and is shown in the Direction column. The PValue provides evidence for whether the majority of genes in the set are DE in the specified direction, whereas the PValue.Mixed tests for differential expression in any direction. FDRs are computed from the corresponding p-values across all sets.

> ## Challenge {.challenge}
>
> 1. Test whether the MYC signalling pathways tend to be differentially expressed between luminal pregnant vs lactating (the second contrast).
> 1. Look for gene sets containing "WNT" in the name and see whether they tend to be differentially expressed in basal pregnant vs lactating.
>

Notes

* A common application of ROAST is to use a set of DE genes that was defined from an analysis of an independent data set. ROAST can then determine whether similar changes are observed in the contrast of interest for the current data set.
* Even for GO-defined gene sets, goana and ROAST have different behaviours. In goana, the significance of differential expression for a GO term is determined relative to other DE genes that are not annotated with that term. In ROAST, only differential expression for the genes in the set are relevant to the significance of that set and its corresponding term. goana depends on a significance cutoff to choose DE genes, whereas ROAST does not require a cutoff and evaluates all genes in the set.
* ROAST estimates p-values by simulation, so the results may change slightly between runs. More precise p-values can be obtained by increasing the number of rotations, albeit at the cost of increased computational time.
* The smallest p-value that can be reported is 1/(2nrot + 1) where nrot is the number of rotations. This lower bound can be decreased by increasing nrot.
