---
title: "RNA-seq analysis in R - Instructor Materials"
author: "Stephane Ballereau, Mark Dunning, Abbi Edwards, Oscar Rueda, Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_notebook:
    toc: yes
  html_document:
    toc: yes
minutes: 300
layout: page
subtitle: Gene Set Testing for RNA-seq
---

```{r setup, include=FALSE}
library(tidyverse)
```

# GSEA analysis

* rank all genes in the `data set`
* identifying the rank positions of all members of the `gene set`
* calculating an enrichment score (ES)
* permute to generate a statistic

## `fgsea`

The `fgsea` package [@Sergushichev2016] implements the same algorithm in R [vignette](http://www.bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html) "fast preranked gene set enrichment analysis (GSEA)".

```{r fgsea}
library(fgsea)
```

```{r loadData}
load("Robjects/Annotated_Results_LvV.RData")
```

## Create ranks

```{r preparedata}
gseaDat <- filter(shrinkLvV, !is.na(Entrez))

ranks <- gseaDat$logFC
names(ranks) <- gseaDat$Entrez
head(ranks)
```

Plot the ranked fold changes.

```{r}
barplot(sort(ranks, decreasing = T))
```

## Load pathways

```{r}
load("Robjects/mouse_H_v5.RData")
pathwaysH <- Mm.H
```

## Conduct analysis

```{r}
fgseaRes <- fgsea(pathwaysH, ranks, minSize=15, maxSize = 500, nperm=1000)
```

* Lets look at the top 10 results.

```{r}
fgseaRes %>% 
    arrange(desc(abs(NES))) %>% 
    top_n(10, -padj)
```

## Enrichment score plot

```{r}
plotEnrichment(pathwaysH[["HALLMARK_ESTROGEN_RESPONSE_EARLY"]], ranks)
```

## GSEA table plot

The function `plotGseaTable` allows us to plot a summary figue showing the 
results for multiple pathways.

```{r}
topPathways <- fgseaRes %>% 
    top_n(20, wt=-padj) %>% 
    arrange(-NES) %>% 
    pull(pathway)

plotGseaTable(pathwaysH[topPathways], 
              rankData, 
              fgseaRes, 
              gseaParam = 0.5)
```

> ## Challenge 1 {.challenge}
>
> Another common way to rank the genes is to order by pvalue, but also, sorting
> so that upregulated genes are at start and downregulated at the other - 
> you can do this combining the sign of the fold change and the pvalue.  
> 1. Rank the genes by statisical significance - you will need to create
> a new ranking value using `-log10({p value}) * sign({Fold Change})`  
> 2. Load the "C2" pathways from the the `data/mouse_c2_v5.RData` file   
> 3. Run `fgsea` using the new ranked genes and the C2 pathways  
> 4. Run `fgsea` using the new ranked genes and the H pathways. How do these 
> results differ from the ones we got when ranking by the fold change alone?  

```{r solution1}
ranks <- -log10(gseaDat$pvalue) * sign(gseaDat$logFC)
names(ranks) <- gseaDat$Entrez  

load("../Robjects/mouse_c2_v5.RData")
pathwaysC2 <- Mm.c2

fgseaResC2 <- fgsea(pathwaysC2, ranks, minSize=15, maxSize = 500, nperm=1000)
head(fgseaResC2[order(padj, -abs(NES)), ], n=10)

fgseaResH <- fgsea(pathwaysH, ranks, minSize=15, maxSize = 500, nperm=1000)
head(fgseaResH[order(padj, -abs(NES)), ], n=20)
```

# GO enrichment analysis

## `goseq`

* GOseq accounts for the gene length bias

```{r goSeqPackage, message=FALSE}
library(goseq)
supportedOrganisms() %>% filter(str_detect(Genome, "mm"))
```

## Create a list of differentially expressed genes

```{r getDEGs}
isSigGene <- shrinkLvV$FDR < 0.01 & !is.na(shrinkLvV$FDR)
genes <- as.integer(isSigGene)
names(genes) <- shrinkLvV$GeneID
```

## Fit the Probability Weighting Function (PWF)

```{r pwFunction}
pwf <- nullp(genes, "mm10", "ensGene", bias.data = shrinkLvV$medianTxLength)
```

## Conduct GO enrichment analysis

```{r runGoseq, message=FALSE}
goResults <- goseq(pwf, "mm10","ensGene", test.cats=c("GO:BP"))
```

## Plot the top 10

```{r plotGO}
goResults %>% 
    top_n(10, wt=-over_represented_pvalue) %>% 
    mutate(hitsPerc=numDEInCat*100/numInCat) %>% 
    ggplot(aes(x=hitsPerc, 
               y=term, 
               colour=over_represented_pvalue, 
               size=numDEInCat)) +
        geom_point() +
        expand_limits(x=0) +
        labs(x="Hits (%)", y="GO term", colour="p value", size="Count")
```

## Get the GO information for the GO accessions

```{r getGOinfo}
library(GO.db)
GOTERM[[goResults$category[1]]]
```

> ## Challenge 2 {.challenge}
>
> 1. Create a list of genes that are statistically significant at FDR < 0.01 and
> are up-regulated by at least 4x (logFC>2)
> in lactating mice
> 2. Run a `goseq` analysis on this gene list
> 3. Plot the results
> 4. How is this result different to the previous GO analysis?

```{r solution2, eval=F}
isSigGeneUp <- shrinkLvV$FDR < 0.01 &
    !is.na(shrinkLvV$FDR) &
    shrinkLvV$logFC > 2

genesUp <- as.integer(isSigGeneUp)
names(genesUp) <- shrinkLvV$GeneID

pwf <- nullp(genesUp, "mm10", "ensGene", bias.data = shrinkLvV$medianTxLength)

goResultsUp <- goseq(pwf, "mm10","ensGene", test.cats=c("GO:BP"))

goResultsUp %>%
    top_n(10, wt=-over_represented_pvalue) %>%
    mutate(hitsPerc=numDEInCat*100/numInCat) %>% 
    ggplot(aes(x=hitsPerc, 
               y=term, 
               colour=over_represented_pvalue, 
               size=numDEInCat)) +
        geom_point() +
        expand_limits(x=0) +
        labs(x="Hits (%)", y="GO term", colour="p value", size="Count")
```

# KEGG pathway enrichment analysis

## `clusterProfiler`

* `clusterprofiler` direct online access of the current KEGG database

```{r loadClusterProfiler, message=FALSE}
library(clusterProfiler)
search_kegg_organism('mmu', by='kegg_code')
```

## KEGG enrichment analysis

```{r enrichKEGG}
sigGenes <- shrinkLvV$Entrez[ shrinkLvV$FDR < 0.01 & 
                              !is.na(shrinkLvV$FDR) &
                              abs(shrinkLvV$logFC) > 1 ]
sigGenes <- na.exclude(sigGenes)
kk <- enrichKEGG(gene = sigGenes, organism = 'mmu')
head(kk, n=10)
```

## Visualise a pathway

### In a browser

```{r browseKegg}
browseKEGG(kk, 'mmu03320')
```

### As a file

```{r pathview}
library(pathview)

logFC <- annotLvV$logFC
names(logFC) <- annotLvV$Entrez
pathview(gene.data = logFC, 
         pathway.id = "mmu03320", 
         species = "mmu", 
         limit = list(gene=5, cpd=1))
```


> ## Challenge 3 {.challenge}
>
> 1. Use `pathview` to export a figure for "mmu04060", but this time only
> use genes that are statistically significant at FDR < 0.01

```{r solution3, eval=F}
sigGenes <- shrinkLvV$FDR < 0.01 & !is.na(shrinkLvV$FDR)

logFC <- annotLvV$logFC[sigGenes]
names(logFC) <- annotLvV$Entrez[sigGenes]

pathview(gene.data = logFC, 
         pathway.id = "mmu04060", 
         species = "mmu", 
         limit = list(gene=5, cpd=1))
```
